Machine learning topological phases in real space

### 2020-07-03 8_prb_submission/1_reply_referee/article_with_comments/article.tex
Writing section emergent information entropy wave functions
Move comparisons of accuracy scores to the numerical experiments section

\textcolor{blue}{\citequote{sufficient_accuracy}, \citequote{superior_comparable} We use the term "sufficient accuracy" to denote a system that performs considerably better at predicting the phases of Hamiltonians than the baseline system that simply guesses the most frequently occurring class for all Hamiltonians. This gauge is needed because it ensures that the resulting entropy signatures are meaningful in the sense that they encode a summary of where useful information about the topological phase of a system can be found. We have added a paragraph to the section \textbf{Numerical experiments} to emphasize the use of this baseline. As to "high accuracy", we note that the accuracy scores obtained by us are comparable to the highest scores reported in Phys. Rev. Lett. 120, 066401, where neural networks were trained on wavevector space data in a similar supervised learning task. This second gauge further reinforces that the entropy signatures presented in the article indeed express a realistic summary of the availability of topological information from each lattice site. A small paragraph was added to the section \textbf{Discussion} to explicitly note this point. In another paragraph added to the section \textbf{Discussion}, we further draw comparisons between our paper and the just recently published article Phys. Rev. Research 2, 023283,  whose theme is model interpretability of single-layer feedforward neural networks trained to recognize phase transitions of some topological condensed matter systems. When contrasting our paper to the latter, we emphasize the important distinction between "model interpretability" and "model explainability". While both techniques rely on model introspection, the first is aimed at interpreting the model itself vis-à-vis previously known concepts and properties related to the physics underlying the data set, while in the latter the goal is to use these model introspection tools to arrive at new concepts and ideas about the physical phenomena underlying the data set, which is the main purpose of our work.   }

@@

\vspace{0.25cm}
\textcolor{blue}{\citequote{unsupervised_learning}, \citequote{unsupervised_approaches} Our main goal in the article is to investigate the localizability of topological information along lattices of SSH systems via model explainability tools in machine learning. We developed a pipeline that accomplishes this task in a very straightforward way by performing supervised learning with decision tree-based algorithms, since their use of Shannon entropy as loss function provides a clear way to measure feature importance. This is compounded by the fact that Shannon entropy plays a very important role in information theory and physics, as is evidenced by the entropic uncertainty relations included in this new version of the article. It should be noted, however, that the eigenvector ensembling algorithm can equally be combined with unsupervised learning algorithms as well. This point is now explicitly mentioned in this new version of the paper, in a new paragraph added to the section \textbf{Discussion}. Finally, as we mentioned in our reply to \citequote{pipeline_not_work}, our procedure can be used for example as a powerful dimensionality reduction tool in data-driven physics, which by itself is an unsupervised learning task and can serve as a preprocessing step to further applications of machine learning to physics.}

@@

\citequote{physical_insights} We have added the new section \textbf{Emergent information entropy wave functions} in which we interpret an information entropy signature as a probability distribution arising from an emergent information entropy wave function in the continuum limit of an infinit lattice. Therefore, the paper now performs a full cycle of data-driven physics, as we propose a method to investigate topological phase transitions using real space eigenvector data from simulations of SSH systems (the eigenvector ensembling algorithm), use model explainability to extract new, previously unknown properties of the SSH systems (the information entropy signatures) and finally interpret these new properties in the light of theoretical physics. It is important to point that the majority of physics papers published to date on the use of machine learning in the investigation of condensed matter systems have been proofs of concept aimed at showing that the machine learning algorithms that have become ubiquitously famous during the last decade indeed are able to recognize the known relevant patterns associated with specific physical phenomena. Our paper goes beyond this task as we use machine learning and model explainability to discover new properties of the physical systems under consideration and (with the addition of this new section) map these previously unknown properties into new theoretical  constructs.

@@

\vspace{0.25cm}
\citequote{experiments_1_2} We have added references to the relevant subsections whenever each numerical experiment is mentioned outside of its subsection throughout the text. The designation "experiment 1" is used to abbreviate the subsection \textbf{Experiment 1: Learning a first-neighbor hopping SSH model with decision trees}. Similarly, the designation "experiment 2" is used to abbreviate the subsection \textbf{Experiment 2: Learning a first- and second-neighbor hoppings SSH model with random forests}.

@@

\vspace{0.25cm}
\citequote{evaluation_metrics}

@@

\vspace{0.25cm}
\citequote{inf_entropy_sig}

@@

\textcolor{blue}{As performance metrics, we report here both accuracy of predicted classes for eigenvectors as well as accuracy of predicted classes for Hamiltonians obtained from eigenvector ensembling. These accuracy scores are to be gauged against the baseline of a system that simply guesses the most frequent class for all eigenvectors (or Hamiltonians). Checking against this baseline is important because it indicates whether the decision trees are in fact learning the underlying patterns that relate real space coordinates to winding numbers, and therefore whether the associated information entropy signature is meaningful or not.}\citequote{sufficient_accuracy} \citequote{evaluation_metrics}

@@

Figures \ref{figexp1_exp} and \ref{figexp2_exp} respectively illustrate single iterations of experiment 1 (section \ref{experiment_1}) and experiment 2 (section \ref{experiment_2}) as seen from parameter space. The accuracy statistics presented in the following subsections and probability heatmaps shown in figures \ref{ssh1_heatmaps} and \ref{ssh2_heatmaps} were obtained after bootstrapping each experiment $n_{exp}$ = 100 times. The recovered probability heatmaps faithfully portray the phase diagrams in figure \ref{fig:phasediagrams}, with clear phase transition lines appearing in the regions of highest uncertainty.

@@

\begin{equation}
\label{entropy_density}
F_S(x) = \int_{0}^x\rho_S(x')dx'
\end{equation}
where $x \in [0,1]$ is a spatial coordinate along the lattice and $\rho_S(x)$ is the local Shannon information entropy density function in the continuum limit.

\textcolor{blue}{
Given the quantum nature of the phase transitions being discussed, the information entropy density function $\rho_S(x)$ can be naturally interpreted as the squared magnitude of a spatial Shannon information entropy wave function,
}
\begin{equation}
\label{shannon_wave_function}
\rho_S(x) = |\psi_S(x)|^2,
\end{equation}
the amount of topological information available from a single point in the continuous 1D string then being expressed in bra-ket notation by Born's rule,

@@

\textcolor{blue}{
The information entropy density function $\rho_S$ devised in this section furnishes a physics-grounded interpretation of the information entropy signatures discovered in the previous section from sheer data analysis of finite SSH systems in real space. In particular, the uncertainty relations \eqref{heisenberg} and \eqref{hirschman} demonstrate the existence of a trade-off between the localizability of information in topological phase transitions in real space and wavevector space. 
}

@@

As we have seen, the use of real space data enabled us to investigate how topological information is distributed in SSH systems. This was demonstrated by the information entropy signatures of section \ref{information_entropy_signatures}, which were recovered from the Shannon entropy of ensembles of eigenvectors in each numerical experiment and led us to the emergent information entropy wave functions of section \ref{emergent_information_entropy_wave_functions}. The existence of such signals that can be recovered from data from many distinct physical systems but are hard to conceptualize from sheer theoretical reasoning provides a clear example of how machine learning and model explainability can be important tools in the investigation of quantum materials.

>>>>

\textcolor{blue}{\citequote{sufficient_accuracy}, \citequote{superior_comparable} We use the term "sufficient accuracy" to denote a system that performs considerably better at predicting the phases of Hamiltonians than the baseline system that simply guesses the most frequently occurring class for all Hamiltonians. This gauge is needed because it ensures that the resulting entropy signatures are meaningful in the sense that they encode a summary of where useful information about the topological phase of a system can be found. We have added a paragraph to the section \textbf{Numerical experiments} to emphasize the use of this baseline. As to "high accuracy", we note that the accuracy scores obtained by us are comparable to the highest scores reported in Phys. Rev. Lett. 120, 066401, where neural networks were trained on wavevector space data in a similar supervised learning task. This second gauge further reinforces that the entropy signatures presented in the article indeed express a realistic summary of the availability of topological information from each lattice site. A small paragraph was added to the section \textbf{Discussion} to explicitly note this point.}

\textcolor{blue}{\citequote{superior_comparable}In another paragraph added to the section \textbf{Discussion}, we further draw comparisons between our paper and the just recently published article Phys. Rev. Research 2, 023283,  whose theme is model interpretability of single-layer feedforward neural networks trained to recognize phase transitions of some topological condensed matter systems. When contrasting our paper to the latter, we emphasize the important distinction between "model interpretability" and "model explainability". While both techniques rely on model introspection, the first is aimed at interpreting the model itself vis-à-vis previously known concepts and properties related to the physics underlying the data set, while in the latter the goal is to use these model introspection tools to arrive at new concepts and ideas about the physical phenomena underlying the data set, which is the main purpose of our work.   }

@@

\vspace{0.25cm}
\textcolor{blue}{\citequote{unsupervised_learning} Our main goal in the article is to investigate the localizability of topological information along lattices of SSH systems via model explainability tools in machine learning. We developed a pipeline that accomplishes this task in a very straightforward way by performing supervised learning with decision tree-based algorithms, since their use of Shannon entropy as loss function provides a clear way to measure feature importance. This is compounded by the fact that Shannon entropy plays a very important role in information theory and physics, as is evidenced by the new section \textbf{Emergent information entropy wave functions} added to this new version of the article.}

\textcolor{blue}{\citequote{unsupervised_approaches} Furthermore, it should be noted that the eigenvector ensembling algorithm can equally be combined with unsupervised learning algorithms as well. This point is now explicitly mentioned in this new version of the paper, in a new paragraph added to the section \textbf{Discussion}. Finally, as we mentioned in our reply to \citequote{pipeline_not_work}, our procedure can be used for example as a powerful dimensionality reduction tool in data-driven physics, which by itself is an unsupervised learning task and can serve as a preprocessing step to further applications of machine learning to physics.}

@@

\textcolor{blue}{\citequote{physical_insights} We have added the new section \textbf{Emergent information entropy wave functions} in which we interpret the information entropy signature as a probability distribution arising from an emergent information entropy wave function in the continuum limit of an infinit lattice. Therefore, the paper now performs a full cycle of data-driven physics, as we i) propose a method to investigate topological phase transitions using real space eigenvector data from simulations of SSH systems (the eigenvector ensembling algorithm), ii) use model explainability to extract new, previously unknown properties of the finite SSH systems (the information entropy signatures) and finally iii) interpret these new properties in the light of theoretical physics.}

\textcolor{blue}{We find it noteworthy that the majority of physics papers published to date on the use of machine learning in the investigation of condensed matter systems have been proofs of concept aimed at showing that the machine learning algorithms that have become ubiquitously famous during the last decade indeed are able to recognize the relevant patterns associated with previously known physical phenomena. Our paper goes beyond this task as we use machine learning and model explainability to discover new properties from the analysis of data from the physical systems under consideration and (with the addition of the new section \textbf{Emergent information entropy wave functions}) map these previously unknown properties into new theoretical  constructs.}

@@

\vspace{0.25cm}
\textcolor{blue}{\citequote{experiments_1_2} We have added references to the relevant subsections whenever each numerical experiment is mentioned outside of its subsection throughout the text. The designation "experiment 1" is used to abbreviate the subsection \textbf{Experiment 1: Learning a first-neighbor hopping SSH model with decision trees}. Similarly, the designation "experiment 2" is used to abbreviate the subsection \textbf{Experiment 2: Learning a first- and second-neighbor hoppings SSH model with random forests}.}

@@

\vspace{0.25cm}
\textcolor{blue}{\citequote{evaluation_metrics} We have added more detailed explanations about the evaluation metrics and probability heatmaps to the section \textbf{Numerical experiments}.}

@@

\vspace{0.25cm}
\textcolor{blue}{\citequote{inf_entropy_sig} The definition of information entropy signatures is included in the new section \textbf{Emergent information entropy wave functions}, where an extended theoretical interpretation of the results obtained with the eigenvector ensembling algorithm is also presented.}

@@

\textcolor{blue}{As performance metrics, we report here both accuracy of predicted classes for eigenvectors as well as accuracy of predicted classes for Hamiltonians obtained from eigenvector ensembling. These accuracy scores are to be gauged against the baseline of a system that simply guesses the most frequent class for all Hamiltonians. Checking against this baseline is important because it indicates whether the decision trees are in fact learning the underlying patterns that relate real space coordinates to winding numbers, and therefore whether the associated information entropy signature is meaningful or not.}\citequote{sufficient_accuracy} \citequote{evaluation_metrics}

@@

Figures \ref{figexp1_exp} and \ref{figexp2_exp} respectively illustrate single iterations of experiment 1 (section \ref{experiment_1}) and experiment 2 (section \ref{experiment_2}) as seen from parameter space. The accuracy statistics presented in the following subsections, as well as the probability heatmaps and recovered phase diagrams shown in figures \ref{ssh1_heatmaps} and \ref{ssh2_heatmaps} were obtained after bootstrapping each experiment $n_{exp}$ = 100 times. \textcolor{blue}{Thus, each probability heatmap shown in figures \ref{ssh1_heatmap_0}-\ref{ssh1_heatmap_1} and \ref{ssh2_heatmap_-1}-\ref{ssh2_heatmap_2} represents the averaged fraction of eigenvectors of each Hamiltonian in the grid that were classified with a given winding number accross $n_{exp}$ = 100 experiments. The recovered phase diagrams \ref{ssh1_heatmap} and \ref{ssh2_heatmap} are constructed by superposing the correponding probability heatmaps}. As these figure make clear, the recovered phase diagrams faithfully portray the true phase diagrams in figure \ref{fig:phasediagrams}, with clear phase transition lines appearing in the regions of highest uncertainty.

@@

\begin{equation}
\label{entropy_density}
F_S(x) = \int_{0}^x\rho_S(x')dx'
\end{equation}
where $x \in [0,1]$ is a spatial coordinate along the continuous lattice (which we shall call from now on string, to emphasize its continuity as opposed to a discrete lattice) and $\rho_S(x)$ is the local Shannon information entropy density function in the continuum limit.

\textcolor{blue}{
Given the quantum nature of the phase transitions being discussed, the information entropy density function $\rho_S(x)$ can be naturally interpreted as the squared magnitude of a spatial Shannon information entropy wave function,
}
\begin{equation}
\label{shannon_wave_function}
\rho_S(x) = |\psi_S(x)|^2,
\end{equation}
the amount of topological information available from a single point in the string then being expressed in bra-ket notation by Born's rule,

@@

\textcolor{blue}{
The information entropy density function $\rho_S$ devised in this section furnishes a physics-grounded interpretation of the information entropy signatures discovered in the previous section from sheer data analysis of finite SSH systems in real space. In particular, the uncertainty relations \eqref{heisenberg} and \eqref{hirschman} express concisely the trade-off between the localizability of information in topological phase transitions in real space and wavevector space. 
}

@@

As we have seen, the use of real space data enabled us to investigate how topological information is spatially distributed in SSH systems. This was demonstrated by the information entropy signatures of section \ref{information_entropy_signatures}, which were recovered from the Shannon entropy of ensembles of eigenvectors in each numerical experiment and led us to the emergent information entropy wave functions of section \ref{emergent_information_entropy_wave_functions}. The existence of such signals that can be recovered from data from many distinct physical systems but are hard to conceptualize from sheer theoretical reasoning provides a clear example of how machine learning and model explainability can be important tools in the investigation of quantum materials.
